Relational Parametricity for Algebraically Indexed Types

### 2012-06-14 paper/geometry-twocolumn.tex
Changes to Section 2
- Move the information flow and metric space examples to after the
  general framework (now in section-further-examples.tex)

- Write a new introduction to section 2 to highlight the structure of
  the geometry type system

- Small fixes throughout the geometry text

\section{Motivating Examples}
\label{sec:motivating-examples}

We motivate this paper's goal of investigating algebraically-indexed
types and their relational interpretations by presenting several
examples. Our leading example is a novel type system for writing
programs that manipulate geometric data, while maintaining geometric
invariants through the types. The other examples are enforcement of
information flow properties via indexing by a logical theory
(\autoref{sec:information-flow}), and incorporation of metric
properties into types (\autoref{sec:continuity-analysis}).

\subsection{2-Dimensional Geometry and Origin Invariance}

@@

Putting this in concrete terms, if we have two origins, $(0,0)$ and
$(10,20)$, then the tuple $(1,1)$ with respect to the first origin and
$(11,21)$ with respect to the second origin represent the \emph{same}
point. It is merely an artifact of our representation of points as
pairs of numbers that these appear to be different. When we write
programs that manipulate data that represents points, we ought to
ensure that our programs are invariant with respect to changes in the
choice of origin. On the other hand, vectors are not intended to be
regarded as invariant under change of origin. Vectors represent
offsets, or movements, and the vector $(0,0)$ always represents the
zero offset.

Invariance under change of representation immediately recalls
Reynolds' fable concerning two professors teaching the theory of
complex numbers \cite{reynolds83types}. One professor chooses
rectangular coordinates ($x + iy$), while the other chooses polar
coordinates ($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after
learning the basic operations on complex numbers in the two
representations, the two classes can interact because the theory of
complex numbers is invariant with respect to the choice of
representation. Reynolds formalises the idea of invariance with
respect to changes of representation as preservation of relations. If
a binary relation $R$ relates two representations of a concept---for
example the rectangular and polar representations of the same complex
number are related---then a program that manipulates complex numbers
at a level of abstraction above their specific representation should
preserve the relation $R$. For example, let $f$ be a program taking
complex numbers to complex numbers that is intended to respect
invariance under change of representation. Then if $c$ is a complex
number in rectangular form, and $c'$ is a complex number in polar form
such that $R$ relates them, which we write as $(c,c') \in R$, then it
should be the case that $(f(c), f(c')) \in R$.

We now apply Reynolds' relational approach to the geometric
setting. We define a family of binary relations on $\mathbb{R}^2$
which is indexed by changes of origin. Changes of origin are
represented by vectors in $\mathbb{R}^2$, and form a group under
addition. To emphasise the group structure, and the intended meaning
as translations, we write the collection of all change of origin
offsets as $\Transl{2}$. By quantifying over all changes of origin we
will be able to ensure that our programs are invariant with respect to
the particular choice of origin. The members of the
$\Transl{2}$-indexed family of binary relations $\{ R_{\vec{t}}
\subseteq \mathbb{R}^2 \times \mathbb{R}^2 \}_{\vec{t} \in \Transl{2}}$
are defined as follows:
\begin{displaymath}
  R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}
\end{displaymath}

@@

\subsection{A Type System for Origin Invariance}
\label{sec:type-system-geom-intro}

@@

Our language of types also includes the common type constructors for
function types, $A \to B$, sum types $A + B$, unit type
$\tyPrimNm{unit}$ and tuple types $A \times B$. We also assume a
primitive type $\tyPrimNm{real}$, used to represent scalars. As we
noted at the start of the previous section, the central data structure
in geometric applications is the tuple of real numbers for
representing points and vectors. However, we cannot simply express
this as the type $\tyPrimNm{real} \times \tyPrimNm{real}$ because it
will not have the correct relational interpretation. (Two elements of
type $\tyPrimNm{real}$ will be related if and only if they are equal,
and hence two elements of $\tyPrimNm{real} \times \tyPrimNm{real}$
will be related if and only if they are equal, by Reynolds' definition
of the relational interpretation of tuple types.) Therefore, we
introduce a new type parameterised by expressions $e$ of sort
$\SynTransl{2}$ to represent points \emph{and} vectors:
$\tyPrim{vec}{e}$. We have used the name $\tyPrimNm{vec}$, even though
we have taken pains to separate geometric points and vectors, to
recall the computer science notion of vector as a sequence of values
of homogeneous type which has a known length (in this case two).

@@

However, it is possible to give vector addition a more precise type
that shows how it is variant with change of origin:
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: \SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
\end{displaymath}
We can also negate vectors, yielding a vector which points in the
opposite direction. Negation negates translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
\end{displaymath}
With the primitive operations of addition and negation of vectors, we
can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:\SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

@@

\begin{example}\label{ex:uninhabited-type}
  To this point we have emphasised the properties that we can deduce
  about programs from their types, so called ``free theorems''. By
  refining our relational interpretations of types appropriately it is
  also possible to show that certain types are uninhabited. For
  example, the type $\forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t
    + t} \to \tyPrim{vec}{t}$ does not have any inhabitants. We
  demonstrate this in \autoref{sec:types-indexed-abelian-groups}. In
  order to obtain such a result, we require a more sophisticated
  relational interpretation of free indexing variables than the one we
  sketched above.
\end{example}

@@

From basic linear algebra, we know that the vector operations of
addition, negation and scaling, while not invariant under change of
origin, are invariant under change of basis. Following the same
methodology as we did for origin invariance, we can express basis
invariance by means of preservation of relations indexed by changes of
basis. Change of basis is accomplished by applying an invertible
linear map, and the collection of all invertible linear maps on
$\mathbb{R}^2$ forms the group $\GL{2}$. We reflect in our language the
group $\GL{2}$ as a new indexing sort $\SynGL{2}$, which has
(non-abelian) group structure that we will write multiplicatively. We
also extend the $\tyPrimNm{vec}$ type to be indexed by elements of
$\SynGL{2}$ as well as $\SynTransl{2}$. Thus, $\tyPrim{vec}{B,t}$ is a
vector which varies with change of basis $B$ and change of origin
$t$. Formally, the index-erasure and (simplified) relational semantics
of this type are as follows:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}

@@

\begin{example}\label{ex:area-of-triangle-1}
  Using the operations we have defined in this subsection, we can
  compute the area of a triangle. 
% By the type of this function, we can
%   see that the area of a triangle is invariant under orthogonal
%   transformations and arbitrary translations. Such transformations are
%   known as isometries: distance preserving maps.
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
      \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|} \\
      \mathrm{area}\ [B]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  The calculation is performed in several steps, each of which removes
  some of the symmetry described by the type of
  $\mathrm{area}$. First, the two offset vectors $p_2 - p_1$ and $p_3
  - p_1$ are computed. These operations remove the effect of
  translations on the result, in exactly the same way as the type
  isomorphism we introduced in \exref{ex:type-iso}. We then compute
  the cross product of the two vectors to determine the area of the
  parallelogram described by the sides of the triangle. This value has
  type $\tyPrim{real}{\det B}$, and we have removed some of the
  symmetry due to invertible linear maps. This value still varies with
  the sign of the determinant, so we remove this symmetry by applying
  the $\mathrm{abs}$ function to leave us with a value of type
  $\tyPrim{real}{|\det B|}$. Finally, we multiply by the constant
  $\frac{1}{2}$ to recover the area of the triangle, rather than the
  whole parallelogram.

  If we specialise $\mathrm{area}$ to just orthogonal transformations,
  we obtain the following type, as a consequence of the equation
  $|\det (\iota_OO)| = 1$:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall O\mathord:\SynOrth{2}, t\mathord:\SynTransl{2}.\ \\
      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1}
    \end{array}
  \end{displaymath}
  This type demonstrates that the area of a triangle is invariant
  under orthogonal transformations and translations. Such combined
  transformations are known as isometries---distance preserving maps.
\end{example}

@@

Indexing types by scaling factors brings us to the original
inspiration for the current work: Kennedy's interpretation of his
units of measure type system via scaling invariance
\cite{kennedy97relational}. Kennedy demonstrates how interpreting
types in terms of scaling invariance brings the techniques of
dimensional analysis \fixme{ref} to bear on programming. The types we
have assigned to the real number arithmetic operations in
\autoref{fig:real-ops} are exactly the types Kennedy assigns in his
units of measure system, except for the absolute value operation. The
effect of our assigned type is that, semantically, we are indexing by
non-zero scaling factors, where Kennedy indexes by strictly positive
scaling factors. We further examine the typing of the absolute value
operation in \autoref{sec:types-indexed-abelian-groups}.

In our two-dimensional setting, we can make additions to Kennedy's
one-dimensional scaling invariance. We add an operation $\iota_1$
that, semantically, takes scale factors in $\GL{1}$ to invertible
linear maps in $\GL{2}$: i.e.~takes numbers $s$ to matrices $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$.  This
operation satisfies the equation $\det (\iota_1 s) = s^2$, indicating
that if we scale the plane by $s$ in all directions, we scale areas by
$s^2$. This is illustrated by the next example.

\begin{example}\label{ex:area-of-triangle-2}
  Just as we specialised the type of $\mathrm{area}$ function to
  orthogonal transformations in \exref{ex:area-of-triangle-1}, we can
  also specialise $\mathrm{area}$'s type to scaling
  transformations. This yields the type:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall s\mathord:\SynGL{1}, t\mathord:\SynTransl{2}.\ \\
      \hspace{0.8cm} \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{real}{s^2}
    \end{array}
  \end{displaymath}
  As expected, the area of a triangle varies with the square of
  scalings of the plane, and this is reflected in the type.
\end{example}

Purely scaling linear maps, as generated by $\iota_1$, commute with
all other invertible linear maps, and form the centre of the group
$\GL{2}$. Thus we add the equation $\iota_1(s)B = B\iota_1(s)$ to our
calculus.

@@

  In \autoref{sec:types-indexed-abelian-groups} we revisit Kennedy's
  proof and show that if we add square root as a primitive
  operation---with the type above---then it is still not possible to
  construct the cube root function.
\end{example}

\subsection{Logical Information Flow}
\label{sec:information-flow}

An extreme example of invariance under change of representation is the
tracking of information flow through programs. If a program does not
depend upon a particular piece of information, then it will be
invariant under \emph{all} changes of representation of this
information. As thoroughly described by Sabelfeld and Sands
\cite{sabelfeld01per}, information flow can be captured semantically
through the use of partial equivalence relations (PERs). PERs form an
instance of Reynolds' approach to abstraction and representation
independence. Abadi, Banerjee, Heintze and Riecke \cite{abadi99core}
built a \emph{Core Calculus for Dependency}, using a type system based
around a security level indexed monad $T_lA$. Tse and Zdancewic
\cite{tse04translating} showed that it is possible to translate Abadi
et al.'s calculus into System F, translating the monadic type $T_lA$
to $\alpha_l \to A$ for some free type variable $\alpha_l$, and making
direct use of Reynolds' abstraction theorem to prove information flow
properties. We now show how a refined variant of Tse and Zdancewic's
translation can be expressed via algebraically-indexed types.

We assume a single indexing sort $\mathsf{principal}$, intended to
represent (possibly composite) principals in a system. Principals are
combined using the connectives of boolean logic ($\top$, $\bot$,
$\lor$, $\land$, $\lnot$). For the equational theory of principals, we
assume all the axioms of boolean algebra, and semantically we
interpret composite principals as boolean values: either true ($\top$)
or false ($\bot$). We introduce a single primitive type, indexed by
expressions of sort $\mathsf{principal}$, $\tyPrim{T}{e}$, with the
following interpretations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{T}{e}} & \{*\} \\
    \rsem{\tyPrim{T}{e}}{}\rho & \left\{
      \begin{array}{ll}
        \{(*,*)\} & \textrm{if }\sem{e}\rho = \top \\
        \{\}      & \textrm{if }\sem{e}\rho = \bot
      \end{array}
      \right.
  \end{array}
\end{displaymath}
The relational interpretation of $\tyPrim{T}{e}$ forces this type to
only be inhabited when the boolean expression $e$ evaluates to true
under the environment $\rho$. To write programs, we assume the
following pair of primitive operations:
\begin{eqnarray*}
  \mathrm{proj} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1 \land p_2} \to \tyPrim{T}{p_1} \\
  \mathrm{combine} & : & \forall p_1 p_2\mathord:\mathsf{principal}.\ \tyPrim{T}{p_1} \to \tyPrim{T}{p_2} \to \tyPrim{T}{p_1 \land p_2}
\end{eqnarray*}
We can now adapt Tse and Zdancewic's translation of Abadi et al.'s
monadic type to our setting. We define a type synonym, for each type
$A$ and expression $e$ of sort $\mathsf{principal}$:
\begin{displaymath}
  T_eA = \tyPrim{T}{e} \to A
\end{displaymath}
For every principal $p$, we can endow the types $T_p-$ with the
structure of a monad. This is due to the fact that it is an instance
of the ``environment'' (or ``reader'') monad \fixme{ref}.

\begin{example}
  Suppose we have a program with the following type:
  \begin{displaymath}
    \forall p,q\mathord:\mathsf{principal}. \tyPrim{T}{e} \to T_p(\tyPrimNm{unit} + \tyPrimNm{unit}) \to T_q(\tyPrimNm{unit} + \tyPrimNm{unit})
  \end{displaymath}
  where $e$ is some boolean algebra expression involving $p$ and
  $q$. We will show in \autoref{sec:instantiations} that programs with
  this type are non-constant (i.e.~depend on their input) if and only
  if the formula $e \Rightarrow q \Rightarrow p$ is valid in boolean
  logic.
\end{example}

% Go a step further by considering equivalence relations on word32, and
% their lattice structure. Link this to quantative information
% flow. Will also need to include singleton types.

\subsection{Metric Spaces and Simple Continuity Analysis}
\label{sec:continuity-analysis}

Our final motivating example of algebraically indexed types and
representation independence makes use of the metric space structure of
the real numbers. In the geometry example, we used relational
interpretations that related values via geometrical
transformations. An alternative relational interpretation of the real
numbers is to relate values that are close, in the metric space
sense. By indexing types by a measure of closeness, we can actually
state continuity of functions as a typing problem.

We introduce an indexing sort $\mathsf{R}^{>0}$, representing positive
distances between real numbers. Our type of real numbers is indexed by
expressions of sort $\mathsf{R}^{>0}$, and we use the following
index-erasure and relational interpretations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{real}{e}} & \mathbb{R} \\
    \rsem{\tyPrim{real}{e}}{}\rho & \{ (x,x') \sepbar |x-x'| < \sem{e}\rho \}
  \end{array}
\end{displaymath}
Thus, two real numbers are related if they only differ by the positive
real number assigned to $e$. With this interpretation, and a
straightforward interpretation of existential types, we can state the
standard $\epsilon$-$\delta$ definition of continuity as a type:
\begin{displaymath}
  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
\end{displaymath}
Chaudhuri, Gulwani and Lublinerman \cite{chaudhuri10continuity} have
presented a program logic based approach to verifying the continuity
of programs. Algebraically indexed types give us a way of expressing
and verifying continuity properties of functions in a type based way.

% \begin{eqnarray*}
%   \underline{c} & : & \forall \epsilon\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon} \\
%   (+) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
%   (-) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1 + \epsilon_2} \\
%   (*) & : & \forall \epsilon_1, \epsilon_2\mathord:\mathsf{R}^{>0}.\ \tyPrim{real}{\epsilon_1} \to \tyPrim{real}{\epsilon_2} \to \tyPrim{real}{\epsilon_1\epsilon_2}
% \end{eqnarray*}

% \begin{enumerate}
% \item Need to give types to basic operations on real numbers
% \item Need to add existential types to the calculus
% \end{enumerate}

>>>>

\section{Geometry via Algebraically Indexed Types}
\label{sec:motivating-examples}

We motivate this paper's goal of investigating algebraically-indexed
types and their relational interpretations by presenting a novel type
system for writing programs that manipulate two-dimensional geometric
data. Geometry is rich with operations that are invariant under
transformation: affine operations are invariant under change of
origin, vector space operations are invariant under change of basis,
and the Euclidean dot product is invariant under orthogonal changes of
basis. Moreover, some operations are interestingly variant under
transformations: for example, the cross product varies with scalings
of the plane. We incorporate this extra information about our
geometric primitives into the type system by means of algebraically
indexed types.

\subsection{Origin Invariance and Representation Independence}

@@

Putting this in concrete terms, if we have two origins, $(0,0)$ and
$(10,20)$, then the tuple $(1,1)$ with respect to the first origin and
$(11,21)$ with respect to the second origin represent the \emph{same}
point because they have the same displacement from their respective
origins. It is merely an artifact of our representation of points as
pairs of numbers that these appear to be different. When we write
programs that manipulate data that represents points, we ought to
ensure that our programs are invariant with respect to changes in the
choice of origin. On the other hand, vectors are not intended to be
regarded as invariant under change of origin. Vectors represent
offsets and the vector $(0,0)$ always represents the zero offset.

Invariance under change of representation immediately recalls
Reynolds' fable about two professors teaching the theory of complex
numbers \cite{reynolds83types}. One professor chooses rectangular
coordinates ($x + iy$), while the other chooses polar coordinates
($\alpha\cos\theta + i\alpha\sin\theta$). Happily, after learning the
basic operations on complex numbers in the two representations, the
two classes can interact because the theory of complex numbers is
invariant with respect to the choice of representation. Reynolds
formalises the idea of invariance with respect to changes of
representation as preservation of relations. If a binary relation $R$
relates two representations of a concept---for example the rectangular
and polar representations of the same complex number are
related---then a program that manipulates complex numbers at a level
of abstraction above their specific representation should preserve the
relation $R$. For example, let $f$ be a program taking complex numbers
to complex numbers that respects invariance under change of
representation. Then if $c$ is a complex number in rectangular form,
and $c'$ is a complex number in polar form such that $R$ relates them,
which we write as $(c,c') \in R$, then it is the case that $(f(c),
f(c')) \in R$.

We now apply Reynolds' relational approach to the geometric
setting. We define a family of binary relations on $\mathbb{R}^2$
which is indexed by changes of origin. Changes of origin are
represented by vectors in $\mathbb{R}^2$, and form a group under
addition. To emphasise the intended meaning as translations, we write
the collection of all change of origin offsets as $\Transl{2}$. By
quantifying over all changes of origin we will be able to ensure that
our programs are invariant with respect to the particular choice of
origin. The members of the $\Transl{2}$-indexed family of binary
relations $\{ R_{\vec{t}} \subseteq \mathbb{R}^2 \times \mathbb{R}^2
\}_{\vec{t} \in \Transl{2}}$ are defined as follows:
\begin{displaymath}
  R_{\vec{t}} = \{ (\vec{v}, \vec{v'}) \sepbar \vec{v'} = \vec{v} + \vec{t} \}
\end{displaymath}

@@

\subsection{A Type System for Change of Origin Invariance}
\label{sec:type-system-geom-intro}

@@

Our language of types also includes the common type constructors for
function types, $A \to B$, sum types $A + B$, unit type
$\tyPrimNm{unit}$ and tuple types $A \times B$. We also assume a
primitive type $\tyPrimNm{real}$, used to represent scalars. As we
noted at the start of the previous section, the central data structure
in geometric applications is the tuple of real numbers for
representing points and vectors. However, we cannot simply express
this as the type $\tyPrimNm{real} \times \tyPrimNm{real}$ because it
will not have the correct relational interpretation. (Two elements of
type $\tyPrimNm{real}$ will be related if and only if they are equal,
and hence two elements of $\tyPrimNm{real} \times \tyPrimNm{real}$
will be related if and only if they are equal, by Reynolds' definition
of the relational interpretation of tuple types.) Therefore, we
introduce a new type indexed by expressions $e$ of sort
$\SynTransl{2}$ to represent points \emph{and} vectors:
$\tyPrim{vec}{e}$. The indexing argument $e$ represents the possible
amount of displacement by change of origin this point may have been
subjected to. We have used the name $\tyPrimNm{vec}$, even though we
have taken pains to separate geometric points and vectors, to recall
the computer science notion of vector as a sequence of values of
homogeneous type which has a known length (in this case two).

@@

However, it is possible to give vector addition a more precise type
that shows how it is variant with change of origin:
\begin{displaymath}
  (+) : \forall t_1, t_2 \mathord: \SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 + t_2}
\end{displaymath}
Intuitively, this type says that if the first input vector has been
displaced by $t_1$ and the second by $t_2$, then their sum is
displaced by $t_1 + t_2$. We can also negate vectors, yielding a
vector which points in the opposite direction. Negation negates
translation arguments:
\begin{displaymath}
  \mathrm{negate} : \forall t \mathord: \SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{-t}
\end{displaymath}
With the primitive operations of addition and negation of vectors, we
can define the derived operation of subtraction:
\begin{displaymath}
  \begin{array}{l}
    (-) : \forall t_1,t_2 \mathord:\SynTransl{2}.\ \tyPrim{vec}{t_1} \to \tyPrim{vec}{t_2} \to \tyPrim{vec}{t_1 - t_2} \\
    (-)\ [t_1]\ [t_2]\ p_1\ p_2 = p_1 + \mathrm{negate}\ p_2
  \end{array}
\end{displaymath}

@@

\begin{example}\label{ex:uninhabited-type}
  To this point we have emphasised the properties that we can deduce
  about programs from their types, so called ``free theorems''. By
  refining our relational interpretations of types appropriately it is
  also possible to show that certain types are uninhabited. For
  example, the type $\forall t \mathord: \SynTransl{2}.\
  \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ does not have any
  inhabitants. Intuitively, this is because we have no way of removing
  the extra $t$ in $\tyPrim{vec}{t+t}$ using our operations. We
  demonstrate this indefinability result in
  \autoref{sec:types-indexed-abelian-groups}. In order to obtain this
  result, we require a more sophisticated relational interpretation of
  free translation variables than the one we sketched above.
\end{example}

@@

From basic linear algebra, we know that the vector operations of
addition, negation and scaling, while not invariant under change of
origin, are invariant under change of basis. Following the same
methodology as we did for origin invariance, we can express basis
invariance by means of preservation of relations indexed by changes of
basis. Change of basis is accomplished by applying an invertible
linear map, and the collection of all invertible linear maps on
$\mathbb{R}^2$ forms the \emph{General Linear} group $\GL{2}$. We
reflect in our language the group $\GL{2}$ as a new indexing sort
$\SynGL{2}$, which has (non-abelian) group structure that we will
write multiplicatively. We also extend the $\tyPrimNm{vec}$ type to be
indexed by elements of $\SynGL{2}$ as well as $\SynTransl{2}$. Thus,
$\tyPrim{vec}{B,t}$ is a vector which varies with change of basis $B$
and change of origin $t$. Formally, the index-erasure and (simplified)
relational semantics of this type are as follows:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}=\hspace{0.5em}}l}
    \tySem{\tyPrim{vec}{e_B,e_t}} & \mathbb{R}^2 \\
    \rsem{\tyPrim{vec}{e_B,e_t}}{}\rho & \{ (\vec{v},\vec{v'}) \sepbar \vec{v'} = (\sem{e_B}\rho)\vec{v} + \sem{e_t}\rho \}
  \end{array}
\end{displaymath}

@@

\begin{example}\label{ex:area-of-triangle-1}
  Using the operations we have defined in this subsection, we can
  compute the area of a triangle. 
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
      \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{|\det B|} \\
      \mathrm{area}\ [B]\ [t]\ p_1\ p_2\ p_3 = \frac{1}{2} * \mathrm{abs}\ ((p_2 - p_1) \times (p_3 - p_1))
    \end{array}
  \end{displaymath}
  The calculation is performed in several steps, each of which removes
  some of the symmetry described by the type of
  $\mathrm{area}$. First, the two offset vectors $p_2 - p_1$ and $p_3
  - p_1$ are computed. These operations remove the effect of
  translations on the result, in exactly the same way as the type
  isomorphism we introduced in \exref{ex:type-iso}. We then compute
  the cross product of the two vectors to determine the area of the
  parallelogram described by the sides of the triangle. This value has
  type $\tyPrim{real}{\det B}$, and we have removed some of the
  symmetry due to invertible linear maps. This value still varies with
  the sign of the determinant, so we remove this symmetry by applying
  the $\mathrm{abs}$ function to leave us with a value of type
  $\tyPrim{real}{|\det B|}$. Finally, we multiply by the constant
  $\frac{1}{2}$ to recover the area of the triangle, rather than the
  whole parallelogram.

  If we specialise $\mathrm{area}$ to just orthogonal transformations,
  we obtain the following type, as a consequence of assuming the
  equation $|\det (\iota_OO)| = 1$:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall O\mathord:\SynOrth{2}, t\mathord:\SynTransl{2}.\ \\
      \hspace{0.8cm}\tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{vec}{\iota_OO, t} \to \tyPrim{real}{1}
    \end{array}
  \end{displaymath}
  This type demonstrates that the area of a triangle is invariant
  under orthogonal transformations and translations. Such combined
  transformations are isometries---distance preserving maps.
\end{example}

@@

Indexing types by scaling factors brings us to the original
inspiration for the current work: Kennedy's interpretation of his
units of measure type system via scaling invariance
\cite{kennedy97relational}. Kennedy demonstrates how interpreting
types in terms of scaling invariance brings the techniques of
dimensional analysis \fixme{ref} to bear on programming. The types we
have assigned to the real number arithmetic operations in
\autoref{fig:real-ops} are exactly the types Kennedy assigns in his
units of measure system, except for the absolute value operation. The
effect of our assigned type is that, semantically, we are indexing by
non-zero scaling factors, whereas Kennedy indexes by strictly positive
scaling factors. We further examine the typing of the absolute value
operation in \autoref{sec:types-indexed-abelian-groups}.

In our two-dimensional setting, we can make additions to Kennedy's
one-dimensional scaling invariance. We add an operation $\iota_1$
that, semantically, takes scale factors in $\GL{1}$ to invertible
linear maps in $\GL{2}$: i.e.~takes numbers $s$ to matrices $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$.  This
operation satisfies the equation $\det (\iota_1 s) = s^2$, indicating
that if we scale the plane by $s$ in both directions, we scale areas
by $s^2$. This is illustrated by the next example.

\begin{example}\label{ex:area-of-triangle-2}
  Just as we specialised the type of the $\mathrm{area}$ function to
  orthogonal transformations in \exref{ex:area-of-triangle-1}, we can
  also specialise $\mathrm{area}$'s type to scaling
  transformations. This yields the type:
  \begin{displaymath}
    \begin{array}{@{}l}
      \mathrm{area} : \forall s\mathord:\SynGL{1}, t\mathord:\SynTransl{2}.\ \\
      \hspace{0.8cm} \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{vec}{\iota_1s, t} \to \tyPrim{real}{s^2}
    \end{array}
  \end{displaymath}
  As expected, the area of a triangle varies with the square of
  scalings of the plane, and this is reflected in the type.
\end{example}

Linear maps of the form $\left(
  \begin{smallmatrix}s & 0 \\ 0 & s\end{smallmatrix}\right)$, as
generated by $\iota_1$, commute with all other invertible linear
maps. (These maps are all the elements of $\GL{2}$ that commute with
everything, and form the \emph{centre} of $\GL{2}$.) Therefore we
assume $(\iota_1 s)B = B(\iota_1 s)$ to hold.

@@

  In \autoref{sec:types-indexed-abelian-groups} we revisit Kennedy's
  proof and show that if we add square root as a primitive
  operation---with the type above---then it is still not possible to
  construct the cube root function. The indefinability of cube root is
  of interest due to its relevance to the classical problem of
  trisecting an angle by ruler and compass constructions.
\end{example}
